---
title: "Danish Kings Homework Week 10"
author: "Matti Kjeldsen"
date: "07-03-2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r library,echo=FALSE}
library(tidyverse)

read.csv2("data/danish_monarchs_clean.csv")
```


The task here is to load your Danish Monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the kings

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `kings.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the

separator:___;____  

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r}
# FILL IN THE CODE BELOW and review the outputs
kings1 <- read.csv("data/danish_monarchs_clean.csv")
head(kings1)
glimpse(kings1)


kings2 <- read_csv("data/danish_monarchs_clean.csv")
head(kings2)
glimpse(kings2)


kings3 <- read.csv2("data/danish_monarchs_clean.csv")
head(kings3)
glimpse(kings3)


kings4 <- read_csv2( "data/danish_monarchs_clean.csv")
head(kings4)
glimpse(kings4)
```
Answer: 
1. Which of these functions is a `tidyverse` function? Read data with it below into a `kings` object
  
  read_csv2 is a tidyverse function

2. What is the result of running `class()` on the `kings` object created with a tidyverse function.

  The result is "spec_tbl_df" "tbl_df"      "tbl"         "data.frame" 

3. How many columns does the object have when created with these different functions?
  
  ncol(kings) = 14

4. Show the dataset so that we can see how R interprets each column

  I chose to use the function glimpse(), as it gives me the same informationas the function head(), but glimpse shows all collumns at once, making it easier to assess

```{r}
class(kings1)
#data.frame

class(kings2)
#spec.tbl.df  tbl_df  tbl data.frame

class(kings3)
#data.frame

class(kings4)
#spec_tbl_df  tbl_df  tbl data.frame
```

```{r}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

kings <- kings4

class(kings)
#spec_tbl_df  tbl_df  tbl data.frame

ncol(kings)
#14 collumns

head(kings)
#Nicer to look at
glimpse(kings)
#Shows all collumns at once, making it easier to see and comprehend

```


## Calculate the duration of reign for all the kings in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`


Create a new column called  `duration` in the kings dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

I have already calculated the duration of eachs rulers reign, using Excel functions, but we will calucate this in R and insert these calculations and results in the new column, "duration"
```{r}
# My CODE
na.omit(kings4)
kings_duration <- kings %>% 
  mutate(duration=Slut_regeringsaar - Start_regeringsaar)

```

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `kings` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`


```{r}
# YOUR CODE
kings_duration %>% 
  ggplot(aes(x=Start_regeringsaar , y=Periode))+
  geom_point()+
  geom_smooth()+
  theme_bw()
```

## How many and which kings enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `kings` dataset. Display the result and also  count the resulting rows with `count()`

```{r}
# YOUR CODE
# Step 1: Calculate the average duration
mean_duration <- mean(kings$Periode, na.rm = TRUE)

# Step 2: Filter kings with a longer-than-average duration
longer_than_average_kings <- kings %>%
  filter(kings$Periode > mean_duration) %>% 
  select(Navn)

print(longer_than_average_kings)

# Step 3: Count the resulting rows
longer_than_average_kings_count <- longer_than_average_kings %>%
  count()

# Step 4: Display the count
print(longer_than_average_kings_count)

#24 rulers reigned more than 18.92157 years
```


## How many days did the three longest-ruling monarchs rule? 

* Sort kings by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r}
# YOUR CODE
top_3 <- kings_duration %>% 
  arrange(desc(duration)) %>% 
  slice(1:3)

top_3_days <- top_3 %>% 
  mutate(Days = duration * 365 + ifelse(Periode %% 4 == 0 & (Periode %% 100 != 0 | Periode %% 400 == 0), 1, 0))

print(top_3_days)
  
```


## Challenge: Plot the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r}
# YOUR CODE
# Step 1: Create the 'midyear' column
kings <- kings %>%
  mutate(midyear = Start_regeringsaar + (Slut_regeringsaar - Start_regeringsaar) / 2)

glimpse(kings)

# Step 2: Plot the duration of reign with ggplot
ggplot(kings,aes(x=midyear,y=Periode)) + 
  geom_point() + 
  geom_smooth() + 
  labs(
    title = "Duration of Reign of Danish Monarchs Through Time",
    x = "Year (Midpoint of Reign)",
    y = "Duration of Reign (years)"
  ) + 
  theme_bw()
  
```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!


